\documentclass[a4paper,anonymous]{lipics-v2021}
%\documentclass[a4paper]{lipics-v2021}
%%% Options
% a4paper/letterpaper: for A4 paper format/for US-letter"
% UKenglish/USenglish: for british hyphenation rules/for american hyphenation rules
% numberwithinsect: for section-numbered lemmas etc.
% cleveref: for enabling cleveref support
% autoref: for enabling autoref support
% anonymous: for anonymousing the authors (e.g. for double-blind review)
% thm-restate: for enabling thm-restate support
% authorcolumns: for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors)
% pdfa: for producing a PDF according the PDF/A standard

\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Extending to Binders the Global Trace Condition for Cyclic Lambda-Calculus}
%\titlerunning{Extending to Binders the GTC for Cyclic $\lambda$-Calculus} %TODO optional, please use if title is longer than one line

\author
    {Stefano Berardi}
    {Computer Science Department, Turin University, Torino, Italy}
    {} % (Optional) email address
    {} % (Optional) orcid
    {} % (Optional) author-specific funding acknowledgements

\author
    {Ugo de' Liguoro}
    {Computer Science Department, Turin University, Torino, Italy}
    {} % (Optional) email address
    {} % (Optional) orcid
    {} % (Optional) author-specific funding acknowledgements

\author
    {Daisuke Kimura}
    {Department of Information Science, Toho University, Japan}
    {} % (Optional) email address
    {} % (Optional) orcid
    {} % (Optional) author-specific funding acknowledgements

\author
    {Koji Nakazawa}
    {Graduate School of Informatics, Nagoya University, Japan}
    {} % (Optional) email address
    {} % (Optional) orcid
    {} % (Optional) author-specific funding acknowledgements

\author
    {Makoto Tatsuta}
    {National Institute of Informatics/Sokendai, Tokyo, Japan}
    {} % (Optional) email address
    {} % (Optional) orcid
    {} % (Optional) author-specific funding acknowledgements

    
\authorrunning{S.\,Berardi and U.\,de'\,Liguoro and D.\,Kimura and K.\,Nakazawa and M.\,Tatsuta}
% First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Stefano\,Berardi and Ugo\,de'\,Liguoro and Daisuke\,Kimura and Koji\,Nakazawa and Makoto\,Tatsuta}
% Please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{Theory of computation}
% Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm 

\keywords{Cyclic proofs, Cyclic terms, Fixed-point definition, global trace condition, Lambda calculus, Infinitary lambda calculus}

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{} %optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{We thank Prof. Anupam Das for a valuable scientific correspondence about global trace condition for cyclic terms.} %optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%MY PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}
\usepackage[all]{xy}
\usepackage{amssymb,amsmath,graphicx,color}
\usepackage{mymath,proof,latexsym}
\usepackage{xcolor}
\usepackage[pdftex,outline]{contour}
\usepackage{bussproofs}
\usepackage{tikz} % Tikz is required to use proofzilla
\usetikzlibrary{arrows.meta} % additional arrow styles
\usepackage{proofzilla}
%% PROOFZILLA SETTINGS %%%%%%%%%%%
% Flow-related macros 
% Define node style 
\newemptyvertex{node}{}
% Define edge styles: directed/undirected, color, arrow head etc
\defedgetype{flow}{draw=blue}{}
\defedgetype{dirflow}{-Latex,draw=blue}{}
\defedgetype{flowtwo}{draw=red}{}
\defedgetype{dirflowtwo}{-Latex,draw=red}{}
%%%%% HERE CHANGE THE DEFINITION OF EDGES TO CHANGE THE COLOR  AND OTHER ATRIBUTES %%%%%%
\defedgetype{dirflowblue}{-Latex,draw=blue}{}
\defedgetype{dirflowred}{-Latex,draw=red}{}
\definecolor{black}{gray}{0}
\defedgetype{flowblack}{-Latex,draw=black}{}
% Additional, unnecessary macros (terms etc)
%\input{hobn_macros}  
%% END PROOFZILLA SETTINGS %%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
Circular syntaxes for infinitary lambda-terms
were considered for instance by Ariola, Bl\"{o}m, Klop, Kennaway, Sleep, de Vries.
They allow to interpret lambda-terms written in fixed-point style 
as recursive definitions of infinitary terms. 
%In this way, circular syntax provides an 
%interpretation for basic constructions of functional programs, 
%like the letrec or mu of Clemens Grabmayer, Jan Rochel.
A sufficient, decidable and quite broad condition for reducing such 
infinitary lambda-terms to a numeral can be obtained
by adapting the global trace condition (GTC) to simply typed 
circular lambda-terms with conditional operator. 
GTC was originally proposed by Brotherston and Simpson
as a correctness condition for infinitary proofs.
Several people, including Das, Kuperberg, Pinault, Pous,
proved that GTC implies normalization to a numeral:
this result holds for a wide set of reductions and for closed circular lambda-terms 
with the type N of natural numbers. 

These papers considered a proper subset of infinitary lambda-terms, those
obtained from variables and combinators using a possibly infinite number of application.
Namely, these papers consider no \emph{explicit} clauses for lambda-abstraction,
which is instead defined through combinators.
Our contribution is to add explicit clauses for lambda-abstraction
to the definition of GTC, and to check that 
the extended GTC retains the termination properties of the original GTC. 
%Namely, we prove that the extended GTC implies normalization to some numeral, 
%for all closed infinitary terms of type N and
%for all reductions sequences not reducing inside the 
%predecessor clause for the conditional operator.  
\end{abstract}

\input{macro}

% Section 1
\input{section-introduction}


% Section 2
\input{section-infinite-lambda-terms}

% Section 3
\input{section-trace-infinite-lambda-terms}


% Section 4
\input{section-circular-system-CTlambda}


% Section 5
\input{section-subject-reduction-infinite-lambda-terms}


% Section 6
\input{section-finite-safe-reductions}


% Section 7
\input{section-comparison}


% Section 8
\input{section-conclusion}

\bibliography{cyclic}

\input{section-appendix}

\end{document}

